---
output: 
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
---

# Downloading DEMs and delineating watersheds

There are lots of ways we can delineate watersheds. A particularly easy one is to use the Stream Stats package. You may recall that Stream Stats is a USGS webpage where you can delineate to a point on a stream. It doesn't work (hasn't been developed) in every state (e.g., Wyoming). So it is good to know how to delineate with other packages (e.g., Whitebox).

## StreamStats

Let's first do a watershed delineation using StreamStats. [Here](https://streamstats.usgs.gov/ss/) is the link to the StreamStats GUI.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(dataRetrieval)
devtools::install_github("markwh/streamstats")
library(streamstats)
library(mapview)
library(leaflet)
library(sf)
library(elevatr)
library(raster)
library(viridis)
library(ggnewscale)
library(tidyterra)

```

Let's first have a look at some of the things we can do with StreamStats.

The gage is USGS 06047500 Bozeman Creek near Bozeman MT

```{r}

gage_info <- readNWISsite(site = "06047500")

xlocation <- gage_info$dec_long_va
ylocation <- gage_info$dec_lat_va

# Delineate the watershed using library(streamstats)

ws1 <- delineateWatershed(xlocation = xlocation, 
                          ylocation = ylocation, 
                          crs = 4326, 
                          includeparameters = "true", 
                          includeflowtypes = "true")

# Look at the outline using library(leaflet)
leafletWatershed(ws1)

# StreamStats also gives us some useful information about the watershed.
values <- as_tibble(ws1[["parameters"]][["value"]])
parameters <- as_tibble(ws1[["parameters"]][["name"]])
meta <- as_tibble(ws1[["parameters"]][["description"]]) %>% 
  rename("meta" = "value")

# And we can put that into a useful data frame and write_csv. 
watershed_info <- cbind(parameters, values, meta)

write_csv(watershed_info, "waterhsed_info.csv")

```

Now we can use the shapefile we just got from StreamStats to download a DEM. We will start by writing the shapefile. Next, we will load the shapefile. And then use that shapefile to download DEMs using elevatr. 

```{r}
writeShapefile(watershed = ws1, layer = "layer_name", dir = "ws1_shp", what = "boundary")

# Read in the shape file using library(sf)

boundary <- st_read("ws1_shp/layer_name.shp") 

# Look at the boundary using library(mapview)

mapview(boundary)

# Download a DEM using library(elevatr)

elevation_data <- elevatr::get_elev_raster(locations = boundary, z = 9, clip = "locations", crs = '+init=EPSG:4326')

mapview(elevation_data)

# This requires library(raster)

writeRaster(elevation_data, "dem/bzn_creek_watershed.tif", overwrite = TRUE, package = "raster", crs = '+init=EPSG:4326')

load_dem <- raster("dem/bzn_creek_watershed.tif") # The raster is already in our environment as elevation_data, this line is just to show you how to load a raster in R. 
```

Now that we have a raster there are lots of things we can do with it. One is that we can make it into a df and use ggplot to visualize. 

```{r}

elevation_df <- as.data.frame(elevation_data, xy = TRUE)

colnames(elevation_df)[3] <- "elevation"

# remove rows of data frame with one or more NA's,using complete.cases
elevation_df <- elevation_df[complete.cases(elevation_df), ]

ggplot() +
  geom_raster(data = elevation_df, aes(x = x, y = y, fill = elevation)) +
  geom_sf(data = boundary, fill = NA, color = "white") +
  coord_sf() +
  scale_fill_viridis_c() +
  labs(title = "Bozeman Creek Watershed", x = "Longitude", y = "Latitude", fill = "Elevation (meters)")
```
 
Download USGS DEM with terrainr.

```{r}
library(terrainr)
library(sf)

# We could put the coordinates of the gage in and download. We already have the shapefile so we will use that instead. But the commented out code below would do it. 
# location_of_interest <- data.frame(
#   x = xlocation,
#   y = ylocation
# )
# 
# location_of_interest <- st_as_sf(
#   location_of_interest, 
#   coords = c("x", "y"), 
#   crs = 4326
# )
# 
# box_of_interest <- set_bbox_side_length(location_of_interest, 20000)

output_tiles <- get_tiles(boundary,
                          services = c("elevation"),
                          resolution = 10 # pixel side length in meters
                          )

```

Quick plot with raster::

```{r}
raster::plot(raster::raster(output_tiles[["elevation"]][[1]]))
```

Now we can run some Whitebox algorithms like fill sinks, flow accumulation, flow direction, and stream extraction.  

We always start hydro-analysis with filling/breaching sinks. 

```{r}

wbt_breach_depressions_least_cost(
  dem = "dem/output_tiles.tif",
  output = "dem/bzn_crk_breached.tif",
  dist = 10,
  fill = TRUE)

wbt_fill_depressions_wang_and_liu(
  dem = "dem/bzn_crk_breached.tif",
  output = "dem/bzn_filled_breached.tif"
)
```

Next, we create flow accumulation and pointer rasters. 

```{r}
wbt_d8_flow_accumulation(input = "dem/bzn_filled_breached.tif",
                         output = "dem/D8FA.tif")

wbt_d8_pointer(dem = "dem/bzn_filled_breached.tif",
               output = "dem/D8pointer.tif")
```

Now, we set a pour point and make it a shape file.

```{r}
ppoints <- tribble(
          ~Lon, ~Lat,
          xlocation, ylocation
          )

ppointsSP <- SpatialPoints(ppoints, proj4string = CRS("+proj=longlat +datum=WGS84"))

shapefile(ppointsSP, filename = "dem/pourpoints.shp", overwrite = TRUE)
```


Now we can extract the streams like we did last week. 

You can play with the threshold to see how it affects the drainage density. This threshold requires some knowledge of the watershed, climate, etc. 

```{r}
wbt_extract_streams(flow_accum = "dem/D8FA.tif",
                    output = "dem/raster_streams.tif",
                    threshold = 6000)

wbt_jenson_snap_pour_points(pour_pts = "dem/pourpoints.shp",
                            streams = "dem/raster_streams.tif",
                            output = "dem/snappedpp.shp",
                            snap_dist = 0.0010) #careful with this! Know the units of your data
tmap_mode("view")

pp <- shapefile("dem/snappedpp.shp")
streams <- raster("dem/raster_streams.tif")

tm_shape(streams)+
  tm_raster(legend.show = TRUE, palette = "Blues", )+
tm_shape(pp)+
  tm_dots(col = "red")
```

```{r}
wbt_watershed(d8_pntr = "dem/D8pointer.tif",
              pour_pts = "dem/snappedpp.shp",
              output = "dem/bzn_crk_watershed.tif")

ws <- raster("dem/bzn_crk_watershed.tif")

mapview(ws)

```

We have now delineated the watersehd. We can now make a hillshade for visualization.

```{r}

wbt_hillshade(dem = "dem/output_tiles.tif",
              output = "dem/bzn_creek_hillshade.tif",
              azimuth = 315)

hillshade <- raster("dem/bzn_creek_hillshade.tif")

wbt_raster_to_vector_polygons(input = "dem/bzn_creek_hillshade.tif", 
                              output = "dem/bzn_creek_hillshade.shp", 
                              wd = NULL, 
                              verbose_mode = FALSE)


tm_shape(hillshade) +
  tm_raster(style = "cont", palette = "-Greys", legend.show = FALSE)

```

But let's clip that hillshade to the delineated watershed. 

```{r}

clip_dem <- ws * dem
clip_dem_df <- as.data.frame(clip_dem, xy = TRUE)
colnames(clip_dem_df)[3] <- "elevation"

# remove rows of data frame with one or more NA's,using complete.cases
clip_dem_df <- clip_dem_df[complete.cases(clip_dem_df), ]

# do same for hillshade
clip_hillshade <- ws * hillshade
clip_hillshade_df <- as.data.frame(clip_hillshade, xy = TRUE)
colnames(clip_hillshade_df)[3] <- "hillshade"
clip_hillshade_df <- clip_hillshade_df[complete.cases(clip_hillshade_df), ]
```

Make map with ggplot. 

```{r}
ggplot() +
  geom_raster(data = clip_dem_df, aes(x = x, y = y, fill = elevation)) +
  geom_sf(data = boundary, fill = NA, color = "white") +
  coord_sf() +
  scale_fill_viridis_c() +
  labs(title = "Bozeman Creek Watershed", x = "Longitude", y = "Latitude", fill = "Elevation (meters)")
```

Let's add the pour point, hillshade, and streams to the figure.

```{r}
# Make streams a polygon. 
clip_streams <- ws * streams
clip_streams_df <- as.data.frame(clip_streams, xy = TRUE)
colnames(clip_streams_df)[3] <- "elevation"
clip_streams_df <- clip_streams_df[complete.cases(clip_streams_df), ]
writeRaster(clip_streams, "dem/clip_streams.tif")

wbt_raster_streams_to_vector(streams = "dem/clip_streams.tif",
                             d8_pntr = "dem/D8pointer.tif",
                             output = "dem/streams.shp")

streams_shp <- st_read("dem/streams.shp")


```

Make a map with ggplot, add streams and pour point. 

```{r}
ggplot() +
  geom_raster(data = clip_dem_df, aes(x = x, y = y, fill = elevation)) +
  geom_sf(data = streams_shp, color = "blue") +
  geom_point(data = ppoints, aes(x = Lon, y = Lat), color = "red") + 
  coord_sf() +
  scale_fill_viridis_c() +
  labs(title = "Bozeman Creek Watershed", x = "Longitude", y = "Latitude", fill = "Elevation (meters)")

```

Add hillshade. 

```{r}
ggplot() +
  geom_raster(data = clip_hillshade_df,
              aes(x = x, y = y, fill = hillshade),
              show.legend = FALSE) +
  scale_fill_distiller(palette = "Greys") +
  new_scale_fill() +
  geom_raster(data = clip_dem_df,
              aes(x = x, y = y, fill = elevation),
              alpha = .7) +
  scale_fill_distiller(palette = "GnBu") +
  geom_sf(data = streams_shp, color = "blue") +
  geom_point(data = ppoints, aes(x = Lon, y = Lat), color = "red") +
  guides(fill = guide_colorsteps(barwidth = 20,
                                 barheight = .5,
                                 title.position = "right")) +
  labs(fill = "m") +
  coord_sf() +
  theme_void() +
  theme(legend.position = "bottom")
```

Play with colors in. Here change to cold_humid color. 
```{r}
library(ggnewscale)
library(tidyterra)

ggplot() +
  geom_raster(data = clip_hillshade_df,
              aes(x = x, y = y, fill = hillshade),
              show.legend = FALSE) +
  scale_fill_distiller(palette = "Greys") +
  new_scale_fill() +
  geom_raster(data = clip_dem_df,
              aes(x = x, y = y, fill = elevation),
              alpha = .7) +
  scale_fill_cross_blended_c(palette = "cold_humid") +
  geom_sf(data = streams_shp, color = "blue") +
  geom_point(data = ppoints, aes(x = Lon, y = Lat), color = "red") +
  guides(fill = guide_colorsteps(barwidth = 20,
                                 barheight = .5,
                                 title.position = "right")) +
  labs(fill = "m") +
  coord_sf() +
  theme_void() +
  theme(legend.position = "bottom")
```

Here change to viridis color. 
```{r}

ggplot() +
  geom_raster(data = clip_hillshade_df,
              aes(x = x, y = y, fill = hillshade),
              show.legend = FALSE) +
  scale_fill_distiller(palette = "Greys") +
  new_scale_fill() +
  geom_raster(data = clip_dem_df,
              aes(x = x, y = y, fill = elevation),
              alpha = .7) +
  scale_fill_continuous(type = "viridis") +
  geom_sf(data = streams_shp, color = "blue") +
  geom_point(data = ppoints, aes(x = Lon, y = Lat), color = "red") +
  labs(fill = "Elevation (m)") +
  coord_sf() +
  theme_void() +
  theme(legend.position = "bottom")
```


Make a map with tmap.

```{r}
tmap_mode("plot")

tm_shape(clip_hillshade) +
  tm_raster(style = "cont", palette = "-Greys", legend.show = FALSE) +
  tm_scale_bar() +
  tm_shape(clip_dem) +
  tm_raster(legend.show = FALSE, palette = "-viridis", alpha = 0.5) +
  tm_shape(streams) + 
  tm_lines(col = "blue", lwd = 2) +
  tm_scale_bar() +
  tm_shape(pp) +
  tm_dots(col = "red", size = 0.5) +
  tm_layout(title = "Bozeman Creek \nWatershed")

```



